---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The following code loads the file into a dataframe and converts the date field from a string to a date. Note that the code assumes that the source file has been downloaded and unzipped in the working directory.
 
```{r}
activityData <- read.csv("activity.csv", stringsAsFactors = FALSE)
activityData$date <-  as.POSIXct(activityData$date)
```

## What is mean total number of steps taken per day?

The code below calculates the steps per day and produces a histogram of this data.

```{r}
stepsPerDay <- aggregate(steps ~ date, activityData, sum)
hist(stepsPerDay$steps, main="Histogram of steps per day", xlab="Steps")

```

The histogram shows the distribution is roughly bell-shaped with a peak in the range of 10,000 to 15,000 steps.

The code below calculates the mean and median steps per day. Note that the scipen=999 option simply provides a weight in favour of normal numeric presentation rather than scientific notation.

```{r}
meanSteps <- round(mean(stepsPerDay$steps))
medianSteps <- round(median(stepsPerDay$steps))
options(scipen=999)
```

The mean steps per day is `r meanSteps`. The median steps per day is `r medianSteps`.

## What is the average daily activity pattern?

The code below calculates the average steps per interval and produces a time series plot of this data. 

```{r}
avgStepsPerInterval <- aggregate(steps ~ interval, activityData, mean)
plot(avgStepsPerInterval$interval, avgStepsPerInterval$steps, type="l", 
     xlab="Interval", ylab="Average Steps")
```

The plot shows that on average almost no activity occurs during the first 500 intervals of a day or during the last 200. There is a heavy spike in activity around the 800th interval, with moderate activity for the rest of the day.

The code below calculates which interval had the highest number of average steps.

```{r}

maxInterval <- avgStepsPerInterval[which.max(avgStepsPerInterval$steps), ]

```

The highest average number of steps was `r round(maxInterval$steps)` and this was found in interval `r maxInterval$interval`.

## Imputing missing values

The code below calculates the number of missing values in the original dataset.

```{r}
countMissing <- sum(is.na(activityData$steps))
```

The number of missing values in the original dataset is `r countMissing`.

The code below creates a new dataset with the missing values replaced by the mean for the corresponding interval.

```{r}
newSteps <- ifelse(is.na(activityData$steps), 
                   avgStepsPerInterval$steps[
                     match(activityData$interval, avgStepsPerInterval$interval)
                   ], 
                   activityData$steps)
impActData <- transform(activityData, steps = newSteps)
```

The code below calculates the steps per day using the new dataset and produces a histogram of this data.

```{r}
newStepsPerDay <- aggregate(steps ~ date, impActData, sum)
hist(newStepsPerDay$steps, main="Histogram of steps per day using imputed data", 
     xlab="Steps")
```

The new histogram has a higher peak compared to the original histogram, which is to be expected given that missing values have been replaced with a measure of central tendency (i.e. the mean by interval).

The code below is used to calculate the new mean and median steps per day.

```{r}
impMeanSteps <- round(mean(newStepsPerDay$steps))
impMedianSteps <- round(median(newStepsPerDay$steps))
```

The mean steps per day is `r impMeanSteps`. The median steps per day is `r impMedianSteps`.

In comparison to the non-imputed data, the mean has changed by `r impMeanSteps-meanSteps` and the median has changed by `r impMedianSteps-medianSteps`. As such, replacing missing values has not really changed the measures of central tendency. However, it would have changed measures of average dispersion such as the standard deviation.

## Are there differences in activity patterns between weekdays and weekends?

The code below creates a new column in the imputed dataset to identify whether the record relates to a weekday or a weekend.

```{r}
impActData$dayType  <- ifelse(weekdays(impActData$date) %in% 
                                  c("Saturday", "Sunday"),
                              "weekend", 
                              "weekday")
impActData$dayType  <- as.factor(impActData$dayType)
```

The below code calculates the number of average steps for each combination of day type and interval. This summary is then plotted.

```{r warning=FALSE}
stepsPerDayType <- aggregate(steps ~ dayType + interval, impActData, mean)
library(lattice)
with(stepsPerDayType, 
     xyplot(steps ~ interval | dayType, 
            type = "l",      
            main = "Average Steps by Day Type and Interval",
            xlab = "Intervals",
            ylab = "Average Steps"))
```

The plot shows that there is a spike in activity at around the 800th interval on weekdays whilst there is a more even spread of activity on the weekend. 
